---
title: "HW4_DataSciMath"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment=NA)
```
#### Walt Wells - 08.01-08.07.2016

1. Using matrix operations, describe the solutions for the following family of equations:

```{r}
# x + 2y - 3z = 5
# 2x + y - 3z = 13
# -x + y + 2z= -8
M <- matrix(c(1,2,-1,2,1,1,-3,-3,2),3,3)
B <- c(5,13,-8)
M
```

* a. Find the inverse of the above 3x3 (non-augmented) matrix.
* b. Solve for the solution using R.
* c. Modify the 3x3 matrix such that there exists only one non-zero variable in the solution set.

``` {r}
# a) find inverse of M
Inv <- solve(M)
Inv
# test 
M %*% Inv
# b) find the solution
x <- Inv %*% B
x

# c) Modify the 3x3 matrix such that there exists only one non-zero variable in the solution set.

# not sure best way to do programmatically, but if we make one col the solution (or = 1), allows for other cols to be 0. 
M2 <- matrix(c(1,2,-1,5,13,-8,-3,-3,2),3,3)
M2
round(solve(M2) %*% B, 4)
```

2.  Consider the matrix, q=matrix(c(3,1,4,4,3,3,2,3,2),nrow=3). Let b=c(1,4,5).  Use Cramer's rule and R to determine the solution, x, to qx=b, if one exists.  Show all determinants.

``` {r}
q=matrix(c(3,1,4,4,3,3,2,3,2),nrow=3)
b=c(1,4,5)

# Cramer's rule
c1 <- c(3,1,4)
c2 <- c(4,3,3)
c3 <- c(2,3,2)

# calc determinants
D <- det(q)
Dx <- det(cbind(b, c2, c3))
Dy <- det(cbind(c1, b, c3))
Dz <- det(cbind(c1, c2, b))
D; Dx; Dy; Dz

#solve
D1 <- Dx / D
D2 <- Dy / D
D3 <- Dz / D

D1; D2; D3

#test
solve(q) %*% b

```
